Estimating the Financial Impact of Wildfire Smoke on US National parks

### 2020-05-19 Markdown Documents/Manuscript.Rmd
added train/test figure
added CV table

Paragraphs:
1. Public lands are important
2. NPs are specifically high value
3. NPs are under funded. This underfunding prevents them from doing those important jobs
4. NPs get a lot of their funding from visitation. Vis is influenced by a lot of things. One of these things that's particurlarly out of their control is smoke from adjacent forested areas.
5. Smoke is bad for recreation and should decrease vis.
6. Smoke events have been increasing over the last 50 years. Particularly, these increases have been in the west.
7. These smoke events are expected to worsen via climate change over the next 50 years.
8. Logically, smoke should influence visitation to parks. This decrease should affect income. We seek to assess this effect in a systematic way and backcast what the monetary effects might have been.
9. Given the trends we observe, we project out into the future what the effects might be over the next 50 years.

@@

Our predicted fees correlate well with our observed fees that we have
```{r echo = FALSE, cache=TRUE}
xx<-na.omit(FEdat)
 mytheme<- theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                  plot.title = element_text( size=18, color="black",face="bold"),
                  axis.title.x = element_text( size=18),
                  axis.title.y = element_text( size=18),
                  axis.text=(element_text(color="black", size=14)),
                  legend.title = element_text(colour="black", size=18),
                  legend.text = element_text( size = 14))
xx%>%
  ggplot(.,aes(x=SingleVisitorEntryFee,y=PredictedFee))+
  geom_jitter(size=2, alpha=0.3)+
  geom_smooth(method="lm", se=FALSE, colour="red")+
  geom_text(aes(x=5,y=15),size=10,label=paste("cor =",round(cor(xx$PredictedFee,xx$SingleVisitorEntryFee),digits=2)))+
  ggtitle("Predicted vs Observed Single Visitor Entry Fee")+theme_classic()+mytheme+
  theme(plot.title = element_text(hjust = 0.5))+
  scale_y_continuous(name="Predicted Single Visitor Entry Fee",labels = scales::dollar_format(prefix = "$"))+
  scale_x_continuous(name="Observed Single Visitor Entry Fee",labels = scales::dollar_format(prefix = "$"))
```

@@

To account for the linear trends of visitation, we first calculated the trends, then found the residuals of each datapoint from the trendline.
$$
\begin{aligned}
Resid_{y} &= \begin{cases}y_{obs}-y_{trend}\\ y_{trend} \sim \operatorname{Norm}(\mu_{park},\sigma)\\ \mu_{park} = B_{0}+ B_{1}*X_{1}
\end{cases}
\end{aligned}
$$

@@

@@

```{r echo=FALSE, cache=TRUE}
 plot(fit, regex_pars = "stdsmoke UnitCode",
            prob = 0.5, prob_outer = 0.9)
```

@@

```{r echo=FALSE, warning=FALSE, cache=TRUE,message=FALSE}
library(scales)
dat%>%filter(UnitCode =="ROMO")%>%
ggplot(.,aes(x=date))+
  geom_ribbon(aes(ymin=PredNoSmoke5CI, ymax=PredNoSmoke95CI),fill="#1f78b4",alpha=0.2)+
  geom_ribbon(aes(ymin=PredNoSmoke25CI, ymax=PredNoSmoke75CI),fill="#1f78b4",alpha=0.5)+
  geom_line(aes(y=RecreationVisits,color="black"),size=1.25)+
  geom_line(aes(y=PredNoSmoke50CI,color="#1f78b4"),linetype="longdash",size=1.25)+theme_classic()+
  mytheme + scale_y_continuous(name="Recreational Visits", labels = scales::comma)+xlab("Date")+
  scale_color_identity(name = "",
                       breaks = c("black", "#1f78b4"),
                       labels = c("Observed", "Median Estimate"),
                       guide = "legend")+ggtitle("Rocky Mountain National Park")

@@

## 3.4. Future Projections

>>>>

Paragraphs:
1. Public lands are important
2. NPs are specifically high value
3. NPs are under funded. This under funding prevents them from doing those important jobs
4. NPs get a lot of their funding from visitation. Vis is influenced by a lot of things. One of these things that's particularly out of their control is smoke from adjacent forested areas.
5. Smoke is bad for recreation and should decrease vis.
6. Smoke events have been increasing over the last 50 years. Particularly, these increases have been in the west.
7. These smoke events are expected to worsen via climate change over the next 50 years.
8. Logically, smoke should influence visitation to parks. This decrease should affect income. We seek to assess this effect in a systematic way and backcast what the monetary effects might have been.
9. Given the trends we observe, we project out into the future what the effects might be over the next 50 years.

@@

Our predicted fees correlate well with our observed fees that we have
```{r echo = FALSE, cache=TRUE}
xx<-na.omit(FEdat)
 mytheme<- theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                  plot.title = element_text( size=18, color="black",face="bold"),
                  axis.title.x = element_text( size=18),
                  axis.title.y = element_text( size=18),
                  axis.text=(element_text(color="black", size=14)),
                  legend.title = element_text(colour="black", size=18),
                  legend.text = element_text( size = 14))
xx%>%
  ggplot(.,aes(x=SingleVisitorEntryFee,y=PredictedFee))+
  geom_jitter(size=2, alpha=0.3)+
  geom_smooth(method="lm", se=FALSE, colour="black",size=2,linetype="longdash")+
  geom_text(aes(x=5,y=15),size=10,label=paste("cor =",round(cor(xx$PredictedFee,xx$SingleVisitorEntryFee),digits=2)))+
  ggtitle("Predicted vs Observed Single Visitor Entry Fee")+theme_classic()+mytheme+
  theme(plot.title = element_text(hjust = 0.5))+
  scale_y_continuous(name="Predicted Single Visitor Entry Fee",labels = scales::dollar_format(prefix = "$"))+
  scale_x_continuous(name="Observed Single Visitor Entry Fee",labels = scales::dollar_format(prefix = "$"))
```

@@

To account for the linear trends of visitation, we first calculated the trends, then found the residuals of each datapoint from the trend line.
$$
\begin{aligned}
Resid_{y} &= \begin{cases}y_{obs}-y_{trend}\\ y_{trend} \sim \operatorname{Norm}(\mu_{park},\sigma)\\ \mu_{park} = B_{0}+ B_{1}*X_{1}
\end{cases}
\end{aligned}
$$

@@

```{r echo=FALSE, message=FALSE,cache=TRUE}
knitr::include_graphics(path.expand("~/SmokeProject/Train_Test.png"))
```

@@

```{r echo=FALSE, cache=TRUE, message=FALSE,warning=FALSE}
PNames<-rev(unique(dat$ParkName))

pars<-plot(fit, regex_pars = "stdsmoke UnitCode",
     prob = 0.5, prob_outer = 0.9)

mytheme<- theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                plot.title = element_text( size=18, color="black",face="bold"),
                axis.title.x = element_text( size=18),
                axis.title.y = element_text( size=18),
                axis.text=(element_text(color="black", size=14)),
                legend.title = element_text(colour="black", size=18),
                legend.text = element_text( size = 14))
```

```{r echo=FALSE, cache=TRUE, message=FALSE,warning=FALSE, fig.height = 9, fig.width = 12}
pars+ggplot2::ggtitle("Posterior medians with 50% and 90% intervals")+
  ggplot2::scale_y_discrete(name ="Park",labels=PNames)+
  mytheme+scale_x_continuous(name="Parameter Value", labels = scales::comma)
```

@@

```{r echo=FALSE, warning=FALSE, cache=TRUE,message=FALSE, fig.height = 6, fig.width = 12}
library(scales)
dat%>%filter(UnitCode =="ROMO")%>%
ggplot(.,aes(x=date))+
  geom_ribbon(aes(ymin=PredNoSmoke5CI, ymax=PredNoSmoke95CI),fill="#1f78b4",alpha=0.2)+
  geom_ribbon(aes(ymin=PredNoSmoke25CI, ymax=PredNoSmoke75CI),fill="#1f78b4",alpha=0.5)+
  geom_line(aes(y=RecreationVisits,color="black"),size=1.25)+
  geom_line(aes(y=PredNoSmoke50CI,color="#1f78b4"),linetype="longdash",size=1.25)+theme_classic()+
  mytheme + scale_y_continuous(name="Recreational Visits", labels = scales::comma)+xlab("Date")+
  scale_color_identity(name = "",
                       breaks = c("black", "#1f78b4"),
                       labels = c("Observed", "Median Estimate"),
                       guide = "legend")+ggtitle("Rocky Mountain National Park")

@@

```{r echo=FALSE, message=FALSE,cache=TRUE}
library(tidyverse)
library(rstanarm)
#pull out last 5 years
train1<-dat%>%filter(!Year %in% c("2018","2017","2016","2015","2014"))
test1<-dat%>%filter(Year %in% c("2018","2017","2016","2015","2014"))

#fitTrain<-stan_lmer(VisDiff~stdsmoke|UnitCode,data=train1,adapt_delta = 0.99,chains=8,warmup=2000,iter=8000)
#save(fitTrain,file="CVResidualMod.rds")
load("~/SmokeProject/ModelObjects/CVResidualMod.rds")

preds1<-apply(posterior_predict(fitTrain,test1),2,median)
reals1<-test1$VisDiff

## 2013:2009

train2<-dat%>%filter(!Year %in% c("2013","2012","2011","2010","2009"))
test2<-dat%>%filter(Year %in% c("2013","2012","2011","2010","2009"))
#fitTrain2<-stan_lmer(VisDiff~stdsmoke|UnitCode,data=train2,adapt_delta = 0.99,chains=8,warmup=2000,iter=8000)
#save(fitTrain2,file="CVResidualMod2.rds")
load("~/SmokeProject/ModelObjects/CVResidualMod2.rds")

preds2<-apply(posterior_predict(fitTrain2,test2),2,median)
reals2<-test2$VisDiff

# 2008:2004
train3<-dat%>%filter(!Year %in% c("2008","2007","2006","2005","2004"))
test3<-dat%>%filter(Year %in% c("2008","2007","2006","2005","2004"))
#fitTrain3<-stan_lmer(VisDiff~stdsmoke|UnitCode,data=train3,adapt_delta = 0.99,chains=8,warmup=2000,iter=8000)
#save(fitTrain3,file="CVResidualMod3.rds")
load("~/SmokeProject/ModelObjects/CVResidualMod3.rds")

preds3<-apply(posterior_predict(fitTrain3,test3),2,median)
reals3<-test3$VisDiff

# 2003:1999
train4<-dat%>%filter(!Year %in% c("2003","2002","2001","2000","1999"))
test4<-dat%>%filter(Year %in% c("2003","2002","2001","2000","1999"))
#fitTrain4<-stan_lmer(VisDiff~stdsmoke|UnitCode,data=train4,adapt_delta = 0.99,chains=8,warmup=2000,iter=8000)
#save(fitTrain4,file="CVResidualMod4.rds")
load("~/SmokeProject/ModelObjects/CVResidualMod4.rds")

preds4<-apply(posterior_predict(fitTrain4,test4),2,median)
reals4<-test4$VisDiff

# 1998:1994
train5<-dat%>%filter(!Year %in% c("1998","1997","1996","1995","1994"))
test5<-dat%>%filter(Year %in% c("1998","1997","1996","1995","1994"))
#fitTrain5<-stan_lmer(VisDiff~stdsmoke|UnitCode,data=train5,adapt_delta = 0.99,chains=8,warmup=2000,iter=8000)
#save(fitTrain5,file="CVResidualMod5.rds")
load("~/SmokeProject/ModelObjects/CVResidualMod5.rds")

preds5<-apply(posterior_predict(fitTrain5,test5),2,median)
reals5<-test5$VisDiff

# 1993:1989
train6<-dat%>%filter(!Year %in% c("1993","1992","1991","1990","1989"))
test6<-dat%>%filter(Year %in% c("1993","1992","1991","1990","1989"))
#fitTrain6<-stan_lmer(VisDiff~stdsmoke|UnitCode,data=train6,adapt_delta = 0.99,chains=8,warmup=2000,iter=8000)
#save(fitTrain6,file="CVResidualMod6.rds")
load("~/SmokeProject/ModelObjects/CVResidualMod6.rds")

preds6<-apply(posterior_predict(fitTrain6,test6),2,median)
reals6<-test6$VisDiff

# 1988:1980
train7<-dat%>%filter(!Year %in% c("1988","1987","1986","1985","1984","1983","1982","1981","1980"))
test7<-dat%>%filter(Year %in% c("1988","1987","1986","1985","1984","1983","1982","1981","1980"))
#fitTrain7<-stan_lmer(VisDiff~stdsmoke|UnitCode,data=train7,adapt_delta = 0.99,chains=8,warmup=2000,iter=8000)
#save(fitTrain7,file="CVResidualMod7.rds")
load("~/SmokeProject/ModelObjects/CVResidualMod7.rds")

preds7<-apply(posterior_predict(fitTrain7,test7),2,median)
reals7<-test7$VisDiff

#combine them all
preds<-c(preds1,preds2,preds3,preds4,preds5,preds6,preds7)
reals<-c(reals1,reals2,reals3,reals4,reals5,reals6,reals7)

d<-as.data.frame(cbind(as.vector(preds),reals))
names(d)<-c("preds","reals")
mytheme<- theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                plot.title = element_text( size=18, color="black",face="bold"),
                axis.title.x = element_text( size=18),
                axis.title.y = element_text( size=18),
                axis.text=(element_text(color="black", size=14)),
                legend.title = element_text(colour="black", size=18),
                legend.text = element_text( size = 14))
d%>%
  ggplot(.,aes(x=reals,y=preds))+
  geom_point(size=2.5, alpha=0.2)+
  geom_smooth(method="lm", se=FALSE, colour="black",size=2,linetype="longdash")+
  geom_text(aes(x=3e+05,y=-2e+05),size=10,label=paste("cor =",round(cor(d$preds,d$reals),digits=2)))+
  ggtitle("CV Results")+theme_classic()+mytheme+scale_x_continuous("Observed Residual Visitation", labels = scales::comma)+
  scale_y_continuous(name="Predicted Residual Visitation", labels = scales::comma)+
  theme(plot.title = element_text(hjust = 0.5))

```{r echo=FALSE, message=FALSE,cache=TRUE}
x<-data.frame(Model=c(1:7), 
              "TimePeriod"=c("1980 - 1988",
                                         "1989 - 1993",
                                         "1994 - 1998",
                                         "1999 - 2003",
                           "2004 - 2008",
                           "2009 - 2013",
                           "2014 - 2018"), 
              Correlation=c(cor(as.vector(preds1),reals1),
                                                   cor(as.vector(preds2),reals2),
                                                   cor(as.vector(preds3),reals3),
                                                   cor(as.vector(preds4),reals4),
                                                   cor(as.vector(preds5),reals5),
                                                   cor(as.vector(preds6),reals6),
                                                   cor(as.vector(preds7),reals7)))

x$r2<-round(x$Correlation^2,digits=2)
x$Correlation<-round(x$Correlation,digits=2)
names(x)<-c("Model","Time Period", "Correlation", "R^2")
formattable::formattable(x)
```
## 3.4. Future Projections
